RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-10-25 formalframework.tex
removed beliefs, change formalism

We define the following sets:
\begin{itemize}
\item $IETypes = \{Softgoal, Goal, Task, Resource\}$,
\item $Types = IETypes \cup \{Actor, Contr, Dep,$\\ \qquad $Decomp, Belief, GenArg\}$
\item $Names$ is a finite set of strings.
\item $DecompTypes = \{AND,OR,XOR\}$.
\item $ContribValues = \{Break, Hurt, Some\ nega-$ $tive, Unknown, Some\  positive, Help, Make\},$
\end{itemize}
\end{definition}

@@

\begin{definition}[Intentional Element]
\label{def:ie}
An intentional element $ie\in \mathbb{N}\times IETypes\times Names\times DecompType$ is a relation, where $ie = (id, type, name, decomptype)$ means:
\begin{itemize}
\item $id\in \mathbb{N}$ is a unique identifier for the element,
\item $type\in IETypes$ specifies the type of the element,
\item $name \in Names$ is a string description of the element,
\item $decomptype\in DecompType$ refers to the type of decomposition.\footnote{Note that the decomposition type is relevant only if the element is in fact decomposed into other elements, but since the jUCMNav implementation defines the decomposition type on the element, we do the same here.}
\end{itemize}
A set of intentional elements is denoted by $IE$.
\end{definition}

@@

Some of the IEs in Figure~\ref{fig:example-small2} can be formalized using Def.~\ref{def:ie} as follows:
\begin{itemize}
\item $(2, goal, \text{Generate traffic}, XOR)$: This represents the goal ``Generate traffic'', which is XOR-decomposed. Note the identifier ``2'' is not shown in the GRL model, but it is stored internally.
\item $(5, resource, \text{Car objects}, AND)$: Although this element does not decompose into any other elements, it still contains an AND-decomposition. Our motivation for this choice is that it simplifies the formalization of decomposition links. It is in line with with GRL metamodel and the jUCMNav implementation.
\end{itemize}

Using the short-hand notation defined in Def.~\ref{def:notation}, we can make the following statements:
\begin{itemize}
\item $goal_2.name = \text{Generate traffic}$
\item $goal_2.decomptype = XOR$
\item $resource_5.name = \text{Car objects}$
\end{itemize}

@@

\begin{definition}[Belief]
\label{def:belief}
A belief $bel\in \mathbb{N}\times Names$ is a relation where $bel=(id,name)$ means:
\begin{itemize}
\item $id\in\mathbb{N}$ is the identifier of the belief,
\item $name\in Names$ is a string description of the belief.
\end{itemize}
\end{definition}

\paragraph{Rational and example} Beliefs are different from IEs in the sense that they are not being used in decompositions, dependencies, or contributions. They merely provide additional explanation for an element. The only belief in Figure~\ref{fig:example-small2} is formalized as $(1, \text{Not scientifically correct})$.

\begin{definition}[Actor]
\label{def:actor}
An actor $act \in \mathbb{N}\times Names$ is a relation where $act=(id,name)$ means:
\begin{itemize}
\item $id\in\mathbb{N}$ is the identifier of the actor, 
\item $name\in Names$ is a string description of its name.
\end{itemize}

@@

We can formalize the actor of Figure~\ref{fig:example-small2} as $act=(0,\text{Traffic Simulator})$, and we can for instance state $act_0.name = \text{Traffic Simulator}$.

\begin{definition}[Actor-IE Relations]
\label{def:act-ie-relation}
An \emph{Actor-IE relation} $r_{ActIE}\in\mathbb{N}\times\mathbb{N}$ is a relation $(i,j)$ meaning that an actor with id $i$ has intentional element with id $j$.

@@

In Figure~\ref{fig:example-small2}, recall we formalized two IEs and the actor as follows:
\begin{itemize}
\item $(2, goal, \text{Generate traffic}, XOR)$
\item $(5, resource, \text{Car objects}, AND)$
\item $(0, \text{Traffic Simulator})$
\end{itemize}

We can then formalize the Actor-IE relationships for these elements as: $(0,2)$ and $(0,5)$.

@@

\begin{definition}[Belief link]
\label{def:link:belief}
A \emph{belief} link: $bellink\in \mathbb{N}\times\mathbb{N}\times\mathbb{N}$ is a relation such that $bellink = (i,j)$ means
\begin{itemize}
\item $i\in \mathbb{N}$ is the unique identifier of the belief element,
\item $j\in \mathbb{N}$ is the unique identifier of the IE.
\end{itemize}
Intuitively, $bellink = (i,j)$ means that $bel_i$ is a belief for $ie_j$.

A set of belief links is denoted by $Bellink$.
\end{definition}

\paragraph{Rationale and example} The belief links can only be used to connect a belief with an intentional element, and they are not used in any of the algorithms. Instead, they can be understood as metadata, they provide additional documentation on the rationale for making certain design choices. Therefore, belief links do not require unique identifiers.

The fact that belief links only connect beliefs with IEs is not made explicit in the definition above. We make this assumption, together with various assumption, explicit in the definition of a \emph{valid GRL model} (Def.~\ref{def:valid-grl-model}). Figure~\ref{fig:example-small2} contains one belief links. It connects the following two elements:
\begin{itemize}
\item $(1, \text{Not scientifically correct})$,
\item $(4, softgoal, \text{Simple design}, AND)$.
\end{itemize}
The belief link is then formalized as $(1, 4)$.

\begin{definition}[Contribution Link]
\label{def:link:contrib}
A \emph{contribution} link: $contrib\in \mathbb{N}\times\mathbb{N}\times\mathbb{N}\times Contribvalues$ is a relation such that $contrib = (i,j,k,value)$ means
\begin{itemize}
\item $i\in \mathbb{N}$ is the unique identifier of the link,
\item $j\in \mathbb{N}$ is the unique identifier of the IE from which the contribution originates,
\item $k\in \mathbb{N}$ is the unique identifier of the IE to which is contributed.
\item $value\in ContribValues$ is the strength of the contribution.
\end{itemize}
Intuitively, $contrib = (i,j,k,value)$ means that $ie_i$ contributes to $ie_j$ with $value$.

A set of contribution links is denoted by $Contrib$.
\end{definition}

\paragraph{Rationale and example} Similar to IEs, links have identifiers as well. In Fig.~\ref{fig:example-small2}, there are two contribution links. First, let us formalize the related IEs:
\begin{itemize}
\item $(3, softgoal, \text{Realisitc simulation}, AND)$
\item $(4, softgoal, \text{Simple design}, AND)$
\item $(6, task, \text{Create new cars}, AND)$
\end{itemize}

We can formalize the contribution links as $(9, 6, 3, \text{Help})$ and $(10, 6, 4, \text{Hurt})$.

\begin{definition}[Decomposition Link]
\label{def:link:decomp}
A \emph{decomposition} link: $decomp\in \mathbb{N}\times\mathbb{N}\times\mathbb{N}$ is a relation such that $decomp = (i,j,k)$ means
\begin{itemize}
\item $i\in \mathbb{N}$ is the unique identifier of the link,
\item $j\in \mathbb{N}$ is the unique identifier of the decomposing IE,
\item $k\in \mathbb{N}$ is the unique identifier of the IE that is being decomposed.
\end{itemize}
Intuitively, $decomp = (i,j,k)$ means that $ie_i$ decomposes $ie_j$.\footnote{Note that the decomposition type is defined on $IE_{k}$, see Definition~\ref{def:ie}.}

A set of decomposition links is denoted by $Decomp$.
\end{definition}

\paragraph{Rationale and example} Decomposition links are simpler than contribution links, because they only contain information about which elements are being connected. Recall that the decomposition type is stored in the element that is being decomposed. In order to formalize the decomposition links in Figure~\ref{fig:example-small2}, let us first formalize the related IEs:
\begin{itemize}
\item $(5, goal, \text{Generate traffic}, XOR)$
\item $(6, task, \text{Create new cars}, AND)$
\item $(7, task, \text{Keep same cars}, AND)$
\end{itemize}
The decomposition links can be formalized as $(11, 6, 5)$ and $(12, 7, 5)$.

\begin{definition}[Dependency Link]
\label{def:link:dependency}
A \emph{dependency} link: $dep\in \mathbb{N}\times\mathbb{N}\times\mathbb{N}$ is a relation such that $dep = (i,j,k)$ means
\begin{itemize}
\item $i\in \mathbb{N}$ is the unique identifier of the link,
\item $j\in \mathbb{N}$ is the unique identifier of the dependee IE,
\item $k\in \mathbb{N}$ is the unique identifier of the dependent IE.
\end{itemize}
Intuitively, $dep = (i,j,k)$ means that $ie_j$ depends on $ie_j$.

A set of dependency links is denoted by $Dep$.
\end{definition}

\paragraph{Rationale and example} The representation of dependency links is exactly the same as decomposition links. In Figure~\ref{fig:example-small2}, let us formalize the IEs related to the only dependency link:
\begin{itemize}
\item $(2, resource, \text{Car objects}, AND)$
\item $(5, goal, \text{Generate traffic}, XOR)$
\end{itemize}

The dependency link between these two elements is formalized as $(8, 5, 2)$.

\begin{definition}[GRL Model]
\label{def:grl-model}
A \emph{GRL model} $GRL=(IE, Bel, Act, R_{ActIE}, Bellink, Contr, Decomp, Dep)$ consists of:
\begin{itemize}
\item A set $IE$ of intentional elements (Def.~\ref{def:ie}),
\item A set $Bel$ of beliefs (Def.~\ref{def:belief}),
\item A set $Act$ of actors (Def.~\ref{def:actor}),
\item A set $R_{ActIE}$ of Actor-IE relations (Def.~\ref{def:act-ie-relation}),
\item A set $Bellink$ of belief links (Def.~\ref{def:link:belief}),
\item A set $Contr$ of contribution links (Def.~\ref{def:link:contrib}),
\item A set $Decomp$ of decomposition links (Def.~\ref{def:link:decomp}),
\item A set $Dep$ of dependency links (Def.~\ref{def:link:dependency}).
\end{itemize}
\end{definition}

\paragraph{Rationale and example} The definition of a GRL model collects all the previously defined tuples into a single definition. For completeness, we now provide the full specification of Figure~\ref{fig:example-small2}. This model is formalized as
 $GRL=(IE, Bel, Act, R_{ActIE}, Bellink, Contr, Decomp, Dep)$ where:
\begin{flalign*}
&IE=&\{&(2, task, \text{Car objects}, AND),&\\
&   &  &(3, softgoal, \text{Realistic simulation}, AND),&\\
&   &  &(4, softgoal, \text{Simple design}, AND),&\\
&   &  &(5, goal, \text{Generate traffic}, XOR),&\\
&   &  &(6, task, \text{Create new cars}, AND),&\\
&   &  &(7, task, \text{Keep same cars}, AND)\}&\\
&Bel=& &\{(1, \text{Not scientifically correct})\}&\\
&Act=& &\{(0, \text{Traffic Simulator})\}&\\
&R_{ActIE}=& &\{(0,i)\mid 2\le i\le 7\}&\\
&Bellink = & &\{(1, 4)\}&\\
&Contr=& &\{(9, 6, 3, \text{Help}), (10, 6, 4, \text{Hurt})\}&\\
&Decomp=& &\{(11, 6, 5), (12, 7, 5)\}&\\
&Dep=& &\{(8, 5, 2)\}&
\end{flalign*}

\begin{definition}[Valid GRL Model]
\label{def:valid-grl-model}
A GRL model $GRL=(IE, Act, R_{ActIE}, Contr, Decomp, Dep)$ (Def.~\ref{def:grl-model}) is a \emph{valid GRL model} iff the following conditions are satisfied:
\begin{enumerate}
\item ids are globally unique across IEs, Beliefs, Links, and Actors, i.e., let $X,Y\in \{IE,Act, Belief, Contr, Decomp, Dep\}$. For all $X_i$ and $Y_j$: if $i=j$ then $X=Y$ and $X_i=Y_j$.
\item All intentional elements of actors exist: $\forall (i,j)\in R_{ActIE}: act_i \in Act \wedge ie_j \in IE$.
\item An intentional element belongs at most to one actor: $\forall ie_i\in IE: |\{(k,i)\in R_{ActIE}\}| \le 1$.
\item Contribution links connect intentional elements: $\forall (i,j,k,value)\in Contrib: \{ie_j,ie_k\}\subseteq IE$.
\item Decomposition links connect intentional elements: $\forall (i,j,k)\in Decomp: \{ie_j,ie_k\}\subseteq IE$.
\item Dependency links connect intentional elements: $\forall (i,j,k)\in Dep: \{ie_j,ie_k\}\subseteq IE$.
\item Belief links connect beliefs with intentional elements: $\forall (i,j)\in Bellink: bel_i\in Bel\wedge ie_j\in IE$.
\end{enumerate}
\end{definition}

@@

Let us briefly verify that our previous formalization of Figure~\ref{fig:example-small2} satisfies all the constraints of Def.~\ref{def:valid-grl-model}:
\begin{enumerate}
\item All elements in the formalization have different ids, so this constraint is satisfied.
\item $R_{ActIE}$ contains one element $(0,i)$ for each IE with id $i$, so this constraint is satisfied as well. Note that other elements (beliefs and links) are not related to actors. This is in line with the jUCMNav implementation.
\item Since we have only one actor with id 0, and this is the only actor that appears in $R_{ActIE}$, this constraint is satisfied.
\item The contribution links connect elements with ids 3, 4, and 6, which are all IEs.
\item The decomposition links connect elements with ids 5, 6, and 7, which are all IEs.
\item The dependency link connects id 2 with 5, which are both IEs.
\item The belief link connects id 1 with id 4, which are respectively a belief and an IE.
\end{enumerate}

@@

\begin{definition}[Generic Argument]
\label{def:generic-argument}
A generic argument $ga \in \mathbb{N}\times Names$ is a relation such that $ga=(id, name)$ means
\begin{itemize}
\item $id\in \mathbb{N}$ is the identifier of the generic argument,
\item $name\in Names$ is a string description of its name.
\end{itemize}
We may refer to the argument with id $i$ simply with $ga_i$. A set of arguments is denoted by $GA$.
\end{definition}

\paragraph{Rationale and example} A generic counterargument is simply an argument that can be used to attack any previous argument, or an IE. It is different from a belief element, since a generic counterargument is used in the computation to determine whether a RationalGRL element is accepted or rejected. Note that one of the constraints of a GRL model (Def.~\ref{def:grl-model}) is that GRL links (Def.~\ref{def:link:contrib},~\ref{def:link:decomp}, and~\ref{def:link:dependency}) should connect IEs, which means that in GRL generic arguments cannot be connected with GRL links. This is correct, since generic arguments are not part of GRL, so they should also not occur in links.

In Fig.~\ref{fig:example-small3}, there are two generic counterargument which can be formalized as $(19, \text{Redundant})$ and $(20, \text{Necessary})$.

\begin{definition}[Argument]
\label{def:argument}
An argument $A=(elem,type)$ is a pair such that:
\begin{itemize}
\item $elem$ is either an intentional element $ie$ (Def.~\ref{def:ie}), an actor $act$ (Def.~\ref{def:actor}), an Actor-IE relation $r_{ActIE}$ (Def.~\ref{def:act-ie-relation}),  a contribution link $contr$, a decomposition link $decomp$ (Def.~\ref{def:link:decomp}), a dependency link $dep$ (Def.~\ref{def:link:dependency}), or a generic argument $ga$ (Def.~\ref{def:generic-argument}).
\item $type\in Type$ (Def.~\ref{def:set-definitions}) is the type of the argument.
\end{itemize}
\end{definition}

@@

\paragraph{Rationale and example} The attack link is the only link that RationalGRL adds to GRL. In Fig~\ref{fig:example-small3}, there are three attack links. The first two attack links originate from generic counterarguments and involve the following three elements:
\begin{itemize}
\item $A_1 = ((1, Resource, \text{Car objects}, AND), Resource)$
\item $A_{19} = ((19, \text{Necessary}), GenArg)$
\item $A_{20} = ((20, \text{Redundant}), GenArg)$
\end{itemize}

These two attack links are formalized as $(A_{20},A_{19})$ and $(A_{19}, A_1)$.

The third attack link is an attack link created by \emph{replacing} and existing argument (we will discuss this process in more detail in Sect.~\ref{sect:algorithms}). It involves two version of the same IE:
\begin{itemize}
\item $A_{4} = ((5, goal, \text{Generate traffic}, AND), Goal)$
\item $A_{18} = ((5, goal, \text{Generate traffic}, XOR), Goal)$
\end{itemize}

This attack link is formalized at $(A_{18},A_4)$.

@@

\paragraph{Rationale and example} The definition of a RationalGRL model collects all the previously defined tuples into a single definition. For completeness, we now provide the full specification of Figure~\ref{fig:example-small3}. Let us first enumerate all the arguments used in this example:
\begin{flalign*}
A_0 = &((0, Actor, \text{Traffic simulator}), Actor)&\\
A_1 = &((1, task, \text{Car objects}, AND), Task),&\\
A_2 = &((2, softgoal, \text{Real. sim.}, AND), Softgoal),&\\
A_3 = &((3, softgoal, \text{Simple des.}, AND), Softgoal),&\\
A_4 = &((4, goal, \text{Generate traffic}, AND), Goal),&\\
A_5 = &((5, task, \text{Create new cars}, AND), Task),&\\
A_6 = &((6, task, \text{Keep same cars}, AND)\}, Task),&\\
A_7 = &((7, 4, 1), Dep),&\\
A_8 = &((8, 5, 2, \text{Help}), Contr),&\\
A_9 = &((9, 5, 3, \text{Hurt}), Contr),&\\
A_{10} = &((10, 5, 4), Decomp),&\\
A_{11} = &((11, 6, 4), Decomp),&\\
A_i = &((0,i-11),ActIE), \text{for } 12\le i\le 17&\\
A_{18} = &((4, goal, \text{Generate traffic}, XOR), Goal),&\\
A_{19} = &((18, \text{Redundant}), GenArg),&\\
A_{20} = &((19, \text{Necessary}), GenArg),&
\end{flalign*}

This model is then formalized as $RationalGRL=(Arg, Att)$ where:
\begin{flalign*}
Arg = &\{A_0,A_1,\ldots,A_{20}\}\\
Att=&\{(A_{18},A_4), (A_{19},A_1), (A_{20},A_{19})\}
\end{flalign*}

All the arguments and the attack relations of this RationalGRL model are shown in Figure~\ref{fig:example-small4}. Note the arguments for Actor-IE containment (arguments $A_{12}$ to $A_{20}$) have been omitted from this figure for readability. It can be read from the figure that two arguments are currently rejected, namely $A_4$ and $A_{19}$. However, we did not yet make precise how exactly this is computed. We will do so in the following definitions.

@@

\begin{figure}[ht!]
\centering
\begin{tikzpicture}
        \node[minimum size=1cm] (a20) [argNodeIN] at (-3,0) {$A_{20}$};
        \node[minimum size=1cm] (a19) [argNodeIN] at (0,0) {$A_{19}$};
        \node[minimum size=1cm] (a1) [argNodeIN] at (3,0) {$A_1$};

         \path
    (a20) edge [attackLink] (a19)
    (a19) edge [attackLink] (a1);

@@

Let us explain these definitions using the example argumentation framework in Figure~\ref{fig:goalmodeling:arg2}, which is a subset of Figure~\ref{fig:example-small4} containing only arguments $A_1, A_{19}$, and $A_{20}$. In this example, there are two admissible sets: $\{A_{20}\}$ and $\{A_1, A_{20}\}$. In the last admissible set, $A_{20}$ defends $A_1$ against its attacker $A_{19}$. Sets containing both $A_{19}$ and either $A_1$ or $A_{20}$ are not conflict free, and the sets $\{A_1\}$ and $\{A_{19}\}$ do not defend themselves against $A_{19}$ and $A_{20}$, respectively.

@@

In our example from Figure~\ref{fig:goalmodeling:arg2}, there is one preferred extension, namely $\{A_{20},A_1\}$. Returning to our running example of Figure~\ref{fig:example-small3}, this means that the intentional elements \text{Car objects} (task) and the generic counerargument \emph{Necessary} are both \emph{accepted}, while the generic counter argument ``Redundant'' is rejected.

>>>>

We define the following sets:
\begin{itemize}
\item $IETypes = \{Softgoal, Goal, Task, Resource\}$,
\item $LinkTypes = \{PosContr, NegContr, Dep, Decomp\}$\footnote{Note that for decomposition links, we only distinguish between a positive contribution ($PosContr$) and a negative contribution ($NegContr$). The GRL metamodel specifies a total of seven contribution values. We instead have chosen only two here to simplify some of the notation, but it is straightforward to extend this to seven values.},
\item $Types = IETypes \cup LinkTypes\cup\{Actor, GenArg, ActIE\}$,
\item $Names$ is a finite set of strings,
\item $DecompTypes = \{AND,OR,XOR\}$.
\end{itemize}
\end{definition}

@@

\begin{definition}[Intentional Element]
\label{def:ie}
An intentional element $ie\in \mathbb{N}\times IETypes\times Names\times DecompType$ is a tuple, where $ie = (id, type, name, decomptype)$ means:
\begin{itemize}
\item $id\in \mathbb{N}$ is a unique identifier for the element,
\item $type\in IETypes$ specifies the type of the element,
\item $name \in Names$ is a string description of the element,
\item $decomptype\in DecompType$ refers to the type of decomposition.\footnote{Note that the decomposition type is relevant only if the element is in fact decomposed into other elements, but since the jUCMNav implementation defines the decomposition type on the element, we do the same here.}
\end{itemize}
A set of intentional elements is denoted by $IE$.
\end{definition}

@@

Some of the IEs in Figure~\ref{fig:example-small2} can be formalized using Def.~\ref{def:ie} as follows:
\begin{itemize}
\item $(4, Goal, \text{Generate traffic}, XOR)$: This represents the goal ``Generate traffic'', which is XOR-decomposed. Note the identifier ``2'' is not shown in the GRL model, but it is stored internally.
\item $(1, Resource, \text{Car objects}, AND)$: Although this element does not decompose into any other elements, it still contains an AND-decomposition. Our motivation for this choice is that it simplifies the formalization of decomposition links. It is in line with with GRL metamodel and the jUCMNav implementation.
\end{itemize}

Using the short-hand notation defined in Def.~\ref{def:notation}, we can make the following statements:
\begin{itemize}
\item $goal_4.name = \text{Generate traffic}$
\item $goal_4.decomptype = XOR$
\item $resource_1.name = \text{Car objects}$
\end{itemize}

@@

\begin{definition}[Actor]
\label{def:actor}
An actor $act \in \mathbb{N}\times {Actor}\times Names$ is a tuple where $act=(id,Actor, name)$ means:
\begin{itemize}
\item $id\in\mathbb{N}$ is the identifier of the actor, 
\item $Actor$ states that this tuple is an actor,\footnote{While this element may look redundant in this tuple, it will become important when translating between RationalGRL models and GRL models. Otherwise it would not be possible to identify this tuple as an actor.}
\item $name\in Names$ is a string description of its name.
\end{itemize}

@@

We can formalize the actor of Figure~\ref{fig:example-small2} as $act=(0,Actor,\text{Traffic Simulator})$, and we can for instance state $act_0.name = \text{Traffic Simulator}$.

\begin{definition}[Actor-IE Relations]
\label{def:act-ie-relation}
An \emph{Actor-IE relation} $r_{ActIE}\in\mathbb{N}\times\{ActIE\}\times\mathbb{N}\times\mathbb{N}$ is a tuple where $r_{ActIE}=(id,ActIE, i, j)$ means:
\begin{itemize}
\item $id\in\mathbb{N}$ is the identifier of the relation, 
\item $ActIE$ states that this tuple is an Actor-IE relation,
\item $i\in\mathbb{N}$ is an id of an actor,
\item $j\in\mathbb{N}$ is an id of an IE.
\end{itemize}

@@

In Figure~\ref{fig:example-small2}, recall we formalized two IEs and the actor as follows:
\begin{itemize}
\item $(1, Resource, \text{Car objects}, AND)$
\item $(4, Goal, \text{Generate traffic}, XOR)$
\item $(0, \text{Traffic Simulator})$
\end{itemize}

We can then formalize the Actor-IE relationships for these elements as:
\begin{itemize}
\item $(12, ActIE, 0, 1)$
\item $(15, ActIE, 0, 4)$
\end{itemize}

@@

\begin{definition}[GRL Link]
\label{def:link}
A \emph{GRL link}: $link\in \mathbb{N}\times LinkTypes\times \mathbb{N}\times\mathbb{N}$ is a tuple such that $link = (id,type,src,dest)$ means
\begin{itemize}
\item $id\in \mathbb{N}$ is the unique identifier of the link,
\item $type\in LinkTypes$ specifies the type of the link, 
\item $src\in \mathbb{N}$ is the identifier of the IE from which the link starts,
\item $dest\in \mathbb{N}$ is the identifier of the IE at which the link ends.
\end{itemize}

A set of links is denoted by $Link$.
\end{definition}

We also introduce some simplified notation for the links.

\begin{definition}[Notation]
\label{def:notation2}
\begin{itemize}
\item
We can refer to links of a specific type simply by $type_{id}$. For instance, we can abbreviate the link $(0,PosContr,2,3)$ with $posContr_0$. 
\item 
We can refer to a set of positive contribution links, negative contribution links, dependency links, and decomposition link with respectively $PosContr, NegContr, Dep,$, and $Decomp$. 
\item
We can refer to a set of contribution links that can be either positive or negative simply with $Contr$.
\end{itemize}
\end{definition}

\paragraph{Rationale and example} Similar to IEs, links have identifiers as well. In Fig.~\ref{fig:example-small2}, there are five links. First, let us formalize the related IEs:
\begin{itemize}
\item (1, Resource, \text{Car objects}, AND),
\item (2, Softgoal, \text{Realistic simulation}, AND),
\item (3, Softgoal, \text{Simple design}, AND),
\item (4, Goal, \text{Generate traffic}, XOR),
\item (5, Task, \text{Create new cars}, AND),
\item (6, Task, \text{Keep same cars}, AND).
\end{itemize}

We can formalize the links as follows:
\begin{itemize}
\item $(7, Dep, 4, 1)$
\item $(8, PosContr, 5, 2)$
\item $(9, NegContr, 5, 3)$
\item $(10, Decomp, 5, 4)$
\item $(11, Decomp, 6, 4)$
\end{itemize}

Let us next provide some examples of the notation of Definition~\ref{def:notation2}:
\begin{itemize}
\item $dep_7 = (7, Dep, 4, 1)$
\item $Decomp = \{(10, Decomp, 5, 4), (11, Decomp, 6, 4)\}$
\item $PosContr = \{(8, PosContr, 5, 2)\}$
\item $Contr = \{(8, PosContr, 5, 2), (9, NegContr, 5, 3)\}$
\end{itemize}

\begin{definition}[GRL Model]
\label{def:grl-model}
A \emph{GRL model} $GRL=(IE, Act, R_{ActIE}, Link)$ consists of:
\begin{itemize}
\item A set $IE$ of intentional elements (Def.~\ref{def:ie}),
\item A set $Act$ of actors (Def.~\ref{def:actor}),
\item A set $R_{ActIE}$ of Actor-IE relations (Def.~\ref{def:act-ie-relation}),
\item A set $Link$ of GRL links (Def.~\ref{def:link}).
\end{itemize}
\end{definition}

\paragraph{Rationale and example} The definition of a GRL model collects all the previously defined tuples into a single definition. For completeness, we now provide the full specification of Figure~\ref{fig:example-small2}. This model is formalized as $GRL=(IE, Act, R_{ActIE}, Link)$, where:
\begin{flalign*}
&IE=&\{&(1, Resource, \text{Car objects}, AND),&\\
&   &  &(2, Softgoal, \text{Realistic simulation}, AND),&\\
&   &  &(3, Softgoal, \text{Simple design}, AND),&\\
&   &  &(4, Goal, \text{Generate traffic}, XOR),&\\
&   &  &(5, Task, \text{Create new cars}, AND),&\\
&   &  &(6, Task, \text{Keep same cars}, AND)\}&\\
&Act=& &\{(0, Actor, \text{Traffic Simulator})\}&\\
&R_{ActIE}=& &\{(i+11, ActIE, 0,i)\mid 1\le i\le 6\}&\\
&Link=&\{&(7, Dep, 4, 1)&\\
&     & &(8, PosContr, 5, 2)&\\
&     & &(9, NegContr, 5, 3)&\\
&     & &(10, Decomp, 5, 4)&\\
&     & &(11, Decomp, 6, 4)\}&\\
\end{flalign*}

\begin{definition}[Valid GRL Model]
\label{def:valid-grl-model}
A GRL model $GRL=(IE, Act, R_{ActIE}, Contr, Decomp, Dep)$ (Def.~\ref{def:grl-model}) is a \emph{valid GRL model} iff the following conditions are satisfied:
\begin{enumerate}
\item ids are globally unique across IEs, Beliefs, Links, and Actors, i.e., let $X,Y\in \{IE,Act, Belief, Contr, Decomp, Dep\}$. For all $X_i$ and $Y_j$: if $i=j$ then $X=Y$ and $X_i=Y_j$.
\item Intentional elements of actors exist: $\forall (k, ActIE, i,j)\in R_{ActIE}: act_i \in Act \wedge ie_j \in IE$.
\item An intentional element belongs at most to one actor: $\forall ie_i\in IE: |\{(k,ActIE,j,i)\in R_{ActIE}\}| \le 1$.
\item Links connect intentional elements: $\forall (i,type, j,k)\in Link: \{ie_j,ie_k\}\subseteq IE$.
\end{enumerate}
\end{definition}

@@

Let us briefly verify that our previous formalization of Figure~\ref{fig:example-small2} satisfies all the constraints of Def.~\ref{def:valid-grl-model}:
\begin{enumerate}
\item All elements in the formalization have different ids, so this constraint is satisfied.
\item $R_{ActIE}$ contains one element for each IE with id $i$, so this constraint is satisfied as well. Note that links are not related to actors. This is in line with the GRL metamodel.
\item Since we have only one actor with id 0, and this is the only actor that appears in $R_{ActIE}$, this constraint is satisfied.
\item All links connect IEs: the contribution links connect elements with ids 2, 3, and 5, which are all IEs; the decomposition links connect elements with ids 4, 5, and 6, which are all IEs; and the dependency link connects id 1 with 4, which are both IEs.
\end{enumerate}

@@

\begin{definition}[Generic Argument]
\label{def:generic-argument}
A generic argument $ga \in \mathbb{N}\times \{GenArg\}\times Names$ is a tuple such that $ga=(id, GenArg, name)$ means
\begin{itemize}
\item $id\in \mathbb{N}$ is the identifier of the generic argument,
\item $GenArg$ states that the tuple is a generic argument,
\item $name\in Names$ is a string description of its name.
\end{itemize}
We may refer to the argument with id $i$ simply with $ga_i$. A set of arguments is denoted by $GA$.
\end{definition}

\paragraph{Rationale and example} A generic counterargument is simply an argument that can be used to attack any previous argument, or an IE. Note that one of the constraints of a GRL model (Def.~\ref{def:grl-model}) is that GRL links (Def.~\ref{def:link}) should connect IEs, which means that in GRL generic arguments cannot be connected with GRL links. This is desired, since generic arguments are not part of GRL, so they should also not be connected using GRL links.

In Fig.~\ref{fig:example-small3}, there are two generic counterargument which can be formalized as $(18,GenArg, \text{Redundant})$ and $(19, GenArg, \text{Necessary})$.

\begin{definition}[Argument]
\label{def:argument}
An argument $A$ is on of the following tuples:
\begin{itemize}
\item An intentional element (Def.~\ref{def:ie}), 
\item An actor $act$ (Def.~\ref{def:actor}),
\item An Actor-IE relation $r_{ActIE}$ (Def.~\ref{def:act-ie-relation}), 
\item A link $link$ (Def.~\ref{def:link}),
\item A generic argument $ga$ (Def.~\ref{def:generic-argument}).
\end{itemize}
\end{definition}

@@

\paragraph{Rationale and example} The attack link is the only link that RationalGRL adds to GRL. In Fig~\ref{fig:example-small3}, there are three attack links. The first two attack links originate from generic counterarguments and involve the following three arguments:
\begin{itemize}
\item $A_1 = (1, Resource, \text{Car objects}, AND)$
\item $A_{18} = (18, GenArg, \text{Redundant})$
\item $A_{19} = (19, GenArg, \text{Necessary})$
\end{itemize}

These two attack links are formalized as $(A_{18},A_{1})$ and $(A_{19}, A_{18})$.

The third attack link is an attack link created by \emph{replacing} and existing argument (we will discuss this process in more detail in Sect.~\ref{sect:algorithms}). It involves two version of the same IE:
\begin{itemize}
\item $A_{4} = ((4, Goal, \text{Generate traffic}, AND), Goal)$
\item $A_{20} = ((4, Goal, \text{Generate traffic}, XOR), Goal)$
\end{itemize}

This attack link is formalized at $(A_{20},A_4)$.

@@

\paragraph{Rationale and example} The definition of a RationalGRL model collects all the previously defined tuples into a single definition. For completeness, we now provide the full specification of Figure~\ref{fig:example-small3}. Let us first enumerate all the arguments used in this example:
\begin{flalign*}
A_0 = &(0, Actor, \text{Traffic simulator})&\\
A_1 = &(1, Task, \text{Car objects}, AND),&\\
A_2 = &(2, Softgoal, \text{Real. sim.}, AND),&\\
A_3 = &(3, Softgoal, \text{Simple des.}, AND),&\\
A_4 = &(4, Goal, \text{Generate traffic}, AND),&\\
A_5 = &(5, Task, \text{Create new cars}, AND),&\\
A_6 = &(6, Task, \text{Keep same cars}, AND),&\\
A_7 = &(7, Dep, 4, 1),&\\
A_8 = &(8, PosContr, 5, 2),&\\
A_9 = &(9, NegContr, 5, 3),&\\
A_{10} = &(10, Decomp, 5, 4),&\\
A_{11} = &(11, Decomp, 6, 4),&\\
A_i = &(i, ActIE, 0,i+11),ActIE), \text{for } 1\le i\le 6&\\
A_{18} = &(18, GenArg, \text{Redundant}),&\\
A_{19} = &(19, GenArg, \text{Necessary}),&\\
A_{20} = &(4, Goal, \text{Generate traffic}, XOR),&\\
\end{flalign*}

This model is then formalized as $RationalGRL=(Arg, Att)$ where:
\begin{flalign*}
Arg = &\{A_0,A_1,\ldots,A_{20}\}\\
Att=&\{(A_{18},A_1), (A_{19},A_{18}), (A_{20},A_{4})\}
\end{flalign*}

All the arguments and the attack relations of this RationalGRL model are shown in Figure~\ref{fig:example-small4}. Note the arguments for Actor-IE containment (arguments $A_{12}$ to $A_{17}$) have been omitted from this figure for readability. It can be read from the figure that two arguments are currently rejected, namely $A_4$ and $A_{19}$. However, we did not yet make precise how exactly this is computed. We will do so in the following definitions.

@@

\begin{figure}[ht!]
\centering
\begin{tikzpicture}
        \node[minimum size=1cm] (a19) [argNodeIN] at (-3,0) {$A_{19}$};
        \node[minimum size=1cm] (a18) [argNodeIN] at (0,0) {$A_{18}$};
        \node[minimum size=1cm] (a1) [argNodeIN] at (3,0) {$A_1$};

         \path
    (a19) edge [attackLink] (a18)
    (a18) edge [attackLink] (a1);

@@

Let us explain these definitions using the example argumentation framework in Figure~\ref{fig:goalmodeling:arg2}, which is a subset of Figure~\ref{fig:example-small4} containing only arguments $A_1, A_{18}$, and $A_{19}$. In this example, there are two admissible sets: $\{A_{19}\}$ and $\{A_1, A_{19}\}$. In the last admissible set, $A_{19}$ defends $A_1$ against its attacker $A_{18}$. Sets containing both $A_{18}$ and either $A_1$ or $A_{19}$ are not conflict free, and the sets $\{A_1\}$ and $\{A_{18}\}$ do not defend themselves against $A_{18}$ and $A_{19}$, respectively.

@@

In our example from Figure~\ref{fig:goalmodeling:arg2}, there is one preferred extension, namely $\{A_{19},A_1\}$. Returning to our running example of Figure~\ref{fig:example-small3}, this means that the intentional elements \text{Car objects} (task) and the generic counerargument \emph{Necessary} are both \emph{accepted}, while the generic counter argument ``Redundant'' is rejected.
